---
title: "Environments and lists"
output:
  html_document:
    theme: null
    css: mystyle.css
    toc: yes
---

<!--
%\VignetteEngine{knitr::rmarkdown}
%\VignetteIndexEntry{Environments and lists}
-->

```{r echo = FALSE}
options(width = 100)
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r echo = FALSE}
library(rspeed)
library(microbenchmark)

print.microbenchmark <- function(x, ...) {
  mb_summary <- function(x) {
    res <- summary(x, unit="us")
    data.frame(name = res$expr, median = res$median)
  }
  
  print(mb_summary(x))
}

```

## Accessing hashed and unhashed environments


## `ls`

Doing `ls` on a very simple environment is very slow --- on the order of 10 µs:

```{r}
e <- list2env(list(a=1, b=2, c=3, d=4, e=5, f=6), hash = FALSE)
microbenchmark(ls(e))
```

It may come as a surprise that `ls` is much slower than the `names` function, which does the same thing for lists as `ls` does for environments. In this section we'll look at what makes `ls` slow, and ways of speeding it up.


### Arguments to `ls`

If you look at the definition of `ls`, it's apparent that there is a lot of inefficiency, just from looking at the function arguments:

```{r}
ls
#> function (name, pos = -1L, envir = as.environment(pos), all.names = FALSE, pattern) 
#> {
#>     if (!missing(name)) {
#>         pos <- tryCatch(name, error = function(e) e)
#>         if (inherits(pos, "error")) {
#>             name <- substitute(name)
#>             if (!is.character(name)) 
#>                 name <- deparse(name)
#>             warning(gettextf("%s converted to character string", 
#>                 sQuote(name)), domain = NA)
#>             pos <- name
#>         }
#>     }
#>     all.names <- .Internal(ls(envir, all.names))
#>     ...
#> }
```

Normally you just pass in an environment as the first argument, and then the function goes through a lot of contortions to get that argument to be assigned to `envir`, which it passes on to `Internal(ls())`.

It turns out that simply passing the environment as `envir` avoids all that work, and results in a big performance improvement:

```{r}
e <- list2env(list(a=1, b=2, c=3, d=4, e=5, f=6), hash = FALSE)
microbenchmark(
  ls(e),
  ls(envir = e)
)
```

You should always use the `envir` argument if speed matters.

## `ls` vs `names`

As I mentioned earlier, `ls` is much slower than `names`, and we'll compare the two of them here. For `ls`, we'll use hashed and unhashed environments. We'll also compare large and small environments/lists.

```{r}
# Small
l <- list(a=1, b=2, c=3, d=4, e=5, f=6)
e <- list2env(l, hash = FALSE)
h <- list2env(l, hash = TRUE)

# Large - we'll just grab everything from stats
l2 <- as.list(asNamespace('stats'))
e2 <- list2env(l2, hash = FALSE)
h2 <- list2env(l2, hash = TRUE)

length(l2)
```

The small sets have 6 items; the large sets have `1090` items.

```{r}
microbenchmark(
  names(l),
  ls(envir = e, all.names = TRUE),
  ls(envir = h, all.names = TRUE),
  names(l2),
  ls(envir = e2, all.names = TRUE),
  ls(envir = h2, all.names = TRUE),
  unit = "us"
)
```

For small sets, `ls` is about 10x slower than `names`. For large sets, the difference is on the order of 6000x. `names` is barely slowed down by the larger set, whereas `ls()` takes a huge hit from the larger set. To get the keys from the large environment, it takes about 3ms!

As for hashed and unhashed environments: for small sets, hashed environments are faster; for large sets, unhashed environments are slightly faster.

We can throw one more option into the mix: instead of doing `ls()`, we'll try converting it to a list, then run `names` on the list. We'll add this method to the previous tests.

```{r}
microbenchmark(
  names(l),

  ls(envir = e, all.names = TRUE),
  names(as.list.environment(e, all.names = TRUE)),

  ls(envir = h, all.names = TRUE),
  names(as.list.environment(h, all.names = TRUE)),

  names(l2),

  ls(envir = e2, all.names = TRUE),
  names(as.list.environment(e2, all.names = TRUE)),

  ls(envir = h2, all.names = TRUE),
  names(as.list.environment(h2, all.names = TRUE)),

  unit = "us"
)
```

Bizarrely, our roundabout way of getting the names is much, much faster than running `ls()`.

Notes: We used `ls(all.names = TRUE)` so that it behaves like `names`. Also, we used the explicit `as.list.environment` instead of `as.list` to avoid the cost of S3 method dispatch.


## A C replacement for `ls`

One big difference between `names()` and `ls()` is that the latter sorts the results, and this is a very expensive operation. (I'm not sure why sorting is desirable, especially given its cost.)

The rspeed package contains a C function `C_ls`, which is derived from the `R_lsInternal` function, with the sort removed. We'll add tests of `C_ls` to the ones from before:

```{r}
# Need to copy C_ls object out of namespace for this test
C_ls <- rspeed:::C_ls

microbenchmark(
  names(l),

  ls(envir = e, all.names = TRUE),
  names(as.list.environment(e, all.names = TRUE)),
  .Call(C_ls, e),

  ls(envir = h, all.names = TRUE),
  names(as.list.environment(h, all.names = TRUE)),
  .Call(C_ls, h),

  names(l2),

  ls(envir = e2, all.names = TRUE),
  names(as.list.environment(e2, all.names = TRUE)),
  .Call(C_ls, e2),

  ls(envir = h2, all.names = TRUE),
  names(as.list.environment(h2, all.names = TRUE)),
  .Call(C_ls, h2),

  unit = "us"
)
```

`.Call(C_ls)` isn't as fast as `names()` on a list, but it's still much better than a regular `ls()` and somewhat faster than `names(as.list())`.

### Summary of `ls`

If you don't want to use C code, `names(as.list.environment())` is the fastest way, and it's not too much slower than the `C_ls` function in this package.



## Accessing objects in environments and lists

[To be written up]

```{r}
e <- new.env()
l <- list()

# Assignment into environments and lists
microbenchmark(
  e$a <- 1,
  e[["b"]] <- 1,
  .Primitive("[[<-")(e, "c", 1),
  assign("d", 1, envir = e),
  .Internal(assign("e", 1, e, FALSE)),

  l$a <- 1,
  l[["b"]] <- 1,
  l <- .Primitive("[[<-")(l, "c", 1)
)

# Accessing objects in environments and lists
microbenchmark(
  e$a,
  e[["a"]],
  .Primitive("[[")(e, "a"),
  get("a", e, inherits = FALSE),
  get("a", envir = e, inherits = FALSE),
  .Internal(get("a", e, "any", FALSE)),

  l$a,
  l[["a"]],
  .Primitive("[[")(l, "a"),
  .subset2(l, "a")
)
```


## Checking for existence

```{r}
# Environments
e <- new.env()
e$a <- 1

# Test for existence of `a` (which exists), and `c` (which doesn't)
microbenchmark(
  exists('a', e, inherits = FALSE),
  exists('a', envir = e, inherits = FALSE),
  .Internal(exists('a', e, 'any', FALSE)),
  'a' %in% ls(e, all.names = TRUE),
  is.null(e[['a']]),
  is.null(e$a),

  exists('c', e, inherits = FALSE),
  exists('c', envir = e, inherits = FALSE),
  .Internal(exists('c', e, 'any', FALSE)),
  'c' %in% ls(e, all.names = TRUE),
  is.null(e[['c']]),
  is.null(e$c),

  unit = "us"
)


# Lists
l <- list(a=1)
microbenchmark(
  exists('a', l, inherits = FALSE),
  'a' %in% names(l),
  is.null(l[['a']]),
  is.null(l$a),

  exists('c', l, inherits = FALSE),
  'c' %in% names(l),
  is.null(l[['c']]),
  is.null(l$c),

  unit = "us"
)

```
