---
title: "match.arg timings"
output:
  html_document:
    theme: null
    css: mystyle.css
    toc: yes
---

<!--
%\VignetteEngine{knitr::rmarkdown}
%\VignetteIndexEntry{match.arg timings}
-->

```{r echo = FALSE}
options(width = 100)
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r echo = FALSE}
library(microbenchmark)
```

## `match.arg()` timings

How long does it take to call `match.arg()`? Does the time depend on the number of possible arguments, or which argument is matched?

```{r}
plainarg <- function(x) {
  x
}

matcharg2 <- function(x = c("one", "two")) {
  match.arg(x)
}

matcharg3 <- function(x = c("one", "two", "three")) {
  match.arg(x)
}


library(microbenchmark)
microbenchmark(
  plainarg("test"),

  matcharg2(),
  matcharg2("one"),
  matcharg2("two"),

  matcharg3(),
  matcharg3("one"),
  matcharg3("two"),
  matcharg3("three"),

  unit = "us"
)
```

The number of possible values has a negligible effect on speed. Similarly, if a value is provided, which possible value it matches seems to have a negligible effect. However, if _no_ value is passed in, that saves a little bit of time.

Can we make it any faster?

In the definition of `match.arg`, there is some tricky non-standard evaluation code if the second argument, `choices`, is missing, which is usually the case.


```{r eval=FALSE}
match.arg
#> function (arg, choices, several.ok = FALSE) 
#> {
#>     if (missing(choices)) {
#>         formal.args <- formals(sys.function(sys.parent()))
#>         choices <- eval(formal.args[[deparse(substitute(arg))]])
#>     }
#>     ....
#> }
```

What happens if we pass in some values for `choices`?

```{r}
matcharg2_choices <- function(x = c("one", "two")) {
  match.arg(x, choices = c("one", "two"))
}

microbenchmark(
  plainarg("test"),

  matcharg2(),
  matcharg2("one"),

  matcharg2_choices(),
  matcharg2_choices("one"),

  unit = "us"
)
```

This saves a huge amount of time. In the usual case, where `choices` is missing, most of the time is spent in the NSE.


## Appendix

```{r}
sessionInfo()
```
